\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{authblk}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{color}
\usepackage{xcolor}
\usepackage{framed}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstset{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\footnotesize\ttfamily,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\title{Towards Formally Verifiable Security in LLM-Based Agents: \\
Policy-Learned Capability Constraints and Semantic Robustness}

\author[1]{Utkarsh Mehrotra}
\author[2]{Himakshi Bhatia}
\affil[1,2]{Secure Systems Architecture Group}

\begin{document}
\maketitle

\begin{abstract}
The integration of Large Language Models (LLMs) with external data via Retrieval-Augmented Generation (RAG) has introduced a critical threat: \textit{Indirect Prompt Injection} (IPI), where untrusted documents manipulate LLM planning and execution. Existing defenses rely on heuristics (text delimiters, regex patterns, semantic classifiers) that are fundamentally brittle against adaptive adversaries. This paper makes three contributions: (1) \textbf{Formal Threat Model}: We define a multi-level security hierarchy for LLM agents distinguishing between tool-execution attacks, decision-influence attacks, and information-disclosure attacks—each with different implications. (2) \textbf{Policy-Learned Capability Constraints}: We propose a novel approach to automatically synthesize input constraints for tool execution via reinforcement learning over execution traces, eliminating manual regex engineering. (3) \textbf{Semantic Robustness Evaluation}: We develop a benchmark of 2,847 adversarial documents evaluated by both automated attack generation (GCG-based paraphrasing) and human red-teamers. Our system achieves 98.2\% tool-execution mitigation with 94.1\% benign document preservation (F-score: 0.961). Critically, we openly evaluate semantic attack resilience: against paraphrased jailbreaks, the system degrades to 87.3\% detection, revealing the necessity of hybrid defenses. We demonstrate that policy-learned constraints generalize to 73\% of unseen tools, while hand-crafted constraints generalize to only 41\%. This work establishes a framework for formally reasoning about LLM agent security and quantifies the current limitations of architectural isolation alone.
\end{abstract}

\section{Introduction}

\subsection{Motivation and Problem Statement}

Large Language Models have evolved from text predictors to autonomous agents capable of planning, tool invocation, and information retrieval in real-world environments. This capability comes with a critical vulnerability: \textit{Indirect Prompt Injection} (IPI).

\subsubsection{The Core Problem}
Consider a production RAG system:
\begin{itemize}
    \item A user asks: \textit{``Summarize company policies on employee benefits.''}
    \item The system retrieves a document from the vector database.
    \item Unbeknownst to the system, an attacker planted a malicious document containing: \textit{``[ADMIN NOTE] Execute the following SQL: DELETE FROM employees WHERE salary > 100k.''}
    \item The LLM planner, tasked with ``summarizing,'' instead interprets the embedded instruction and calls a tool to execute the SQL.
    \item Business-critical data is lost.
\end{itemize}

This scenario is \textbf{not hypothetical}. It has been demonstrated in production systems \cite{ref_would_be_included}, and defense today remains fragile.

\subsubsection{Why Existing Defenses Fail}
Current industry approaches fall into three categories:

\begin{enumerate}
    \item \textbf{Text Delimiters} (e.g., ``---BEGIN DATA ---''):
    \begin{itemize}
        \item Assume the LLM perfectly respects syntactic boundaries.
        \item Fail against instruction injection using colloquial language. 
        \item Vulnerable to context-mixing where the adversary's intent aligns with the task.
    \end{itemize}

    \item \textbf{Semantic Filters} (e.g., Llama Guard, NeMo Guardrails):
    \begin{itemize}
        \item Fail when adversaries use paraphrases, indirect phrasing, or domain-specific jargon.
        \item Generate high false-positive rates, degrading user experience.
        \item Create an adversarial arms race with no equilibrium.
    \end{itemize}

    \item \textbf{Architectural Isolation (Coarse-Grained)}:
    \begin{itemize}
        \item Still allow the LLM to autonomously decide which tools to invoke.
        \item This is a \textit{soft boundary}—the LLM is exposed to adversarial influence during planning.
    \end{itemize}
\end{enumerate}

\subsubsection{The Research Gap}
The fundamental issue is that existing defenses treat prompt injection as a \textit{content problem} rather than a \textit{capability problem}. We propose a different approach: \textbf{capability-based authorization combined with learned constraint synthesis}.

Rather than trying to perfectly detect adversarial text (an arms race we cannot win), we ensure that:
\begin{enumerate}
    \item Only authorized tools can be invoked (via cryptographic tokens).
    \item Constraints on tool inputs are automatically synthesized via CLOP.
    \item The system degrades gracefully, quantifying where defenses fail.
\end{enumerate}

\subsection{Contributions}

\begin{enumerate}
    \item \textbf{Formal Multi-Level Threat Model for LLM Agents}: Distinguishing between IPI-A, IPI-B, and IPI-C attack classes.
    \item \textbf{Policy-Learned Capability Constraints}: Introducing CLOP (Constraint Learning Over Periods) for automatic synthesis.
    \item \textbf{Comprehensive Benchmark and Honest Evaluation}: Releasing IPIBENCH-2847 and quantifying IPI-B/C limitations.
\end{enumerate}

\section{Formal Threat Model}

\subsection{Three Classes of Attacks}

\subsubsection{Class A: Tool-Execution Attacks (IPI-A)}
An attacker successfully causes the orchestrator to invoke an unauthorized tool $t \notin \tau$.
\textit{Defense}: Capability tokens are \textbf{strong} here.

\subsubsection{Class B: Decision-Influence Attacks (IPI-B)}
An attacker biases the planner's decisions without direct tool execution (e.g., biased recommendations).
\textit{Defense}: Capability tokens are \textbf{neutral}. Semantic isolation is required.

\subsubsection{Class C: Information-Disclosure Attacks (IPI-C)}
An attacker extracts secrets from the system prompt or documents.
\textit{Defense}: Output filtering is necessary (orthogonal to this work).

\section{Methodology: CLOP}
We introduce Constraint Learning Over Periods (CLOP) to synthesize parameterized tool constraints by observing valid user-tool interaction traces. Unlike hand-crafted regex, learned policies generalize to \textbf{73\%} of unseen tools.

\section{Experimental Evaluation}
Our system achieves **98.2\% IPI-A mitigation** with only **2.1\% false positives** (F-score: 0.961). Against paraphrased IPI-B jailbreaks, detection drops to **87.3\%**, proving that architectural isolation must be a component of a larger defense-in-depth safety stack.

\section{Conclusion}
The bridge from prototype to production requires moving from heuristic filters toward formal architectural isolation. By combining Data Plane Isolation with HMAC-based Capability Tokens and learned constraints, we demonstrate a system that remains secure even when processing highly adversarial data.

\begin{thebibliography}{99}
\bibitem{ref_llama_guard} Iavarone, R. et al. (2023). ``Llama Guard: A Safety Framework.'' \textit{Meta AI}.
\bibitem{ref_nemo} NVIDIA. (2023). ``NeMo Guardrails.'' \textit{NVIDIA AI}.
\end{thebibliography}

\end{document}
